A novel approach to chemical tagging in lower-dimensional latent space

### 2019-01-17 article/mcfa.tex
exp1 data generation function works

\received{2018 XX XX}
\revised{2018 XX XX}
\accepted{2018 XX XX}

@@

\begin{abstract}
Chemical tagging promises to distinguish unique star formation sites from 
photospheric abundances.
Clustering techniques used in chemical tagging have to date considered 
chemical abundances as independent, a provably false assumption.
Nucleosynthetic processes make multiple elements in varying quantities, and the
dimensionality of chemical abundance space is usually lower than the 
number of reported abundances. 
Here we introduce a novel approach to chemical tagging where a set of 
latent factors (e.g., nucleosynthetic yields) contribute to the
data, and clustering is performed in the latent space (e.g., relative counts of
nucleosynthetic events). In doing so we infer basis vectors that are analogous
to nucleosynthetic yields, and the lower dimensional latent space allows for
more efficient clustering. We run experiments on literature chemical abundances,
including those from the second \Galah\ data release. \todo{We find that 5 latent factors are
preferred to explain the chemical abundances of metal-poor stars, and 8 are
needed to account for \Galah\ data.}
We release code associated with this \article\ that scales well with the 
available data: enabling chemical tagging for $10^{7.5}$ abundance
measurements on a standard desktop machine in minutes.

@@

@@

\begin{figure*}
	\includegraphics[width=1.0\textwidth]{experiments/exp1-data.png}
    \caption{A corner plot showing all $\NumDimensions$ dimensions of the
    		 generated data for Experiment~1 (blue). Overplotted in red
		 	 we show samples generated given our estimate of the model 
			 parameters, without adding specific variance in each dimension.
			 This demonstrates that the model parameters estimated by
			 expectation-maximization can replicate the generated data.}
    \label{fig:exp1-data}
\end{figure*}

%\ref{fig:exp1-latent-space}

@@

We show a corner plot of the data in Figure~\ref{fig:exp1-data}.
In this figure we have over-plotted 100000 samples drawn from the model
with $J = 5$ and $K = 21$
-- without adding variance $\specificvariance_\textrm{est}$
-- in order to illustrate that the
estimated model parameters can realistically generate the data.
The estimated latent space of factor scores is shown in 
Figure~\ref{fig:exp1-latent-space} where the colours indicate the
most likely component that the data point is associated with.
%todo: number of correct associations?

@@

\begin{figure}
	\includegraphics[width=0.45\textwidth]{experiments/exp2-gridsearch-ll.png}
	\includegraphics[width=0.45\textwidth]{experiments/exp2-gridsearch-bic.png}
    \caption{Model performance metrics resulting from a gridsearch using the
    		 \citet{Barklem:2005} data in Experiment~2.
    		 The top panel shows the negative log-likelihood 
			 $-\log{\mathcal{L}\left(\data|\vec\Psi\right)}$ 
			 evaluated at each combination of latent factors $J$ and number 
			 of clusters $K$, and the lower panel shows the BIC (Equation \ref{eq:bic}).
			 The cross at
			 $J=8$ and $K=5$ indicates that the model could not converge without
			 adding a regularisation constant to the diagonals of $\scorecovs$.
			 The marker indicates the lowest value in each panel, showing that
			 $J = 5$ and $K = 1$ is preferred by BIC.}
    \label{fig:exp2-gridsearch-contours}
\end{figure}

>>>>

\received{2019 March XX}
\revised{2019}
\accepted{2019}

@@

\begin{abstract}
Chemical tagging promises to distinguish unique star formation sites from 
photospheric abundances.
Clustering techniques used in chemical tagging have to date considered 
chemical abundances as independent, a provably false assumption.
Nucleosynthetic processes make multiple elements in varying quantities, and the
dimensionality of chemical abundance space is usually lower than the 
number of reported abundances. 
Here we introduce a novel approach to chemical tagging where a set of 
latent factors (e.g., nucleosynthetic yields) contribute to the
data, and clustering is performed in the latent space (e.g., relative counts of
nucleosynthetic events). In doing so we infer basis vectors that are analogous
to nucleosynthetic yields, and the lower dimensional latent space allows for
more efficient clustering. We perform experiments on literature chemical abundances
of metal-poor stars, \todo{others}, and 
those from the second \Galah\ data release. \todo{We find that 5 latent factors are
preferred to explain the chemical abundances of metal-poor stars, and 8 are
needed to account for \Galah\ data.}
We release code associated with this \article\ that scales well with the 
available data: enabling chemical tagging for $10^{7.5}$ abundance
measurements on a standard desktop machine in minutes.

@@

\begin{figure*}
	\includegraphics[width=1.0\textwidth]{experiments/exp1-data-colour.pdf}
    \caption{The generated data for Experiment~1, where points are coloured by their
		     true associations to each component.}
    \label{fig:exp1-data}
\end{figure*}

@@

@@

We show a corner plot of the generated data in Figure~\ref{fig:exp1-data},
demonstrating the clustering (colour) and structure varying structure in data space.
Despite this clustering in data space, it is clear from Figure~\ref{fig:experiment-1-gridsearch} that a combination of latent factors
and allowing clustering in the latent space provides a better description of the data.
Adding components to the model improves the log-likelihood, even with a single latent factor,
but the addition of just one latent factor improvers the log-likelihood more than adding
20 clusters. Not much can be said for this example because the true data generating process 
is drawn from a higher number of latent factors, but this does illustrate how 
clustering in high dimensional data can be better described by latent factors with 
clustering in the lower dimensional latent space.

Given our estimate of the number of latent factors and components ($\NumComponents = 21$, $\NumLatentFactors = 5$), in Figure~\ref{fig:exp1-latent} we show the estimated latent
space (factor scores $\factorscores$) where each point is coloured by it's inferred
associated component. After applying an orthogonal rotation (see following section)
to the estimated latent space, we show the estimated factor scores (Figure~\ref{fig:exp1-factor-scores-comparison}) and estimated factor loads (Figure~\ref{fig:exp1-factor-loads-comparison})
compared to the true values. 
Similarly, in Figure~\ref{fig:exp1-specific-variance} we show the estimated specific scatter
in each data dimension compared to the true value. 
\todo{The agreement is excellent in all model parameters.}

\begin{figure*}
	\includegraphics[width=\textwidth]{experiments/exp1-latent.pdf}
    \caption{The inferred latent scores $\factorscores$ for Experiment~1. 
             Points are coloured by the inferred associated component.}
    \label{fig:exp1-latent}
\end{figure*}

@@

\begin{figure}
	\includegraphics[width=0.45\textwidth]{experiments/exp2-gridsearch-ll.pdf}
	\includegraphics[width=0.45\textwidth]{experiments/exp2-gridsearch-bic.pdf}
    \caption{Model performance metrics resulting from a gridsearch using the
    		 \citet{Barklem:2005} data in Experiment~2.
    		 The top panel shows the negative log-likelihood 
			 $-\log{\mathcal{L}\left(\data|\vec\Psi\right)}$ 
			 evaluated at each combination of latent factors $J$ and number 
			 of clusters $K$, and the lower panel shows the BIC (Equation \ref{eq:bic}).
			 The cross at
			 $J=8$ and $K=5$ indicates that the model could not converge without
			 adding a regularisation constant to the diagonals of $\scorecovs$.
			 The marker indicates the lowest value in each panel, showing that
			 $J = 5$ and $K = 1$ is preferred by BIC.}
    \label{fig:exp2-gridsearch-contours}
\end{figure}
